Individual foraging specialization and success change across experience in a virtual predator-prey system

### 2025-04-08 manuscript/manuscript.Rmd
on Windows --> updated MS results

---
title: "Individual foraging specialization and success change across experience in a virtual predator-prey system"
output: 
  word_document:
    reference_docx: manuscript-styles.docx
    fig_caption: yes
bibliography: references.bib
csl: chicago-author-date.csl
---

@@

library(data.table)
library(brms)
library(flextable)
library(officer)
library(dplyr)

@@

The capacity of predators to match their hunting tactics to their prey and to optimize their skills at using them over successive encounters is hypothesized to promote individual foraging specialisation. Through this process, predators should be able to capture more prey. Yet, there are limited empirical assessments showing links between past experience, foraging specialization, and hunting success at the individual level, due to the challenges of monitoring direct interactions in the wild. In this study, we used a virtual predator-prey system (the game *Dead by Daylight*) to investigate how individual predator foraging specialization and success developed across repeated interactions with prey. We found that 68% of predators became either increasingly specialized by always moving at a fast pace (specialised cursorial tactic), or flexible by transitioning between slow and fast speeds (generalist ambush tactic). The predators' strategies were partially matched to their prey's speed, suggesting that changes in hunting behaviour were driven by repeated encounters with their prey. Flexible and specialist foragers achieved similar success overall. Hence, our findings suggest that experience may promote behavioural diversification in predator-prey systems.

@@

Recent evidence suggests that the resource fluctuations predators experience over time (e.g., temporal changes in prey behaviour) may be key in shaping individual behavioural specialization and its fitness consequences [@patrick.weimerskirch2014b;@vandenbosch.etal2019;@santoro.etal2019;@manlick.etal2021]. Theory outlines two contrasting scenarios, with outcomes defined by the predator's ability to learn the optimal strategy in response to resource variability. In the first scenario, individuals should gradually specialize in a foraging tactic (e.g., always moving fast) if it enables them to consistently achieve successful prey captures while learning to hunt in environments with predictable (or stable) resources [@tinker.etal2009;@dukas2019]. Such conditions favor specialist hunters, as they require less energy and time to search for, capture, and handle prey [@weimerskirch2007;@woo.etal2008;@potier.etal2015]. For example, habitat specialization under stable resources is associated with reduced foraging effort and increased offspring growth rates in Herring Gulls (*Larus argentatus*) [@vandenbosch.etal2019]. In the second scenario, predators are expected to be more flexible by learning to use a variety of hunting tactics (i.e., generalists) when they hunt in highly variable environments. Indeed, while there can be costs to attempt novel tactics by trial and error, predators can reduce the consequences of uncertainty in environments under fluctuating resources by learning to adjust their tactic to the prey that they encounter [@dall2010;@mathot.etal2012;@stephens1993;@ishii.shimada2010;@kelley.magurran2011]. @szopa-comley.ioannou2022 showed that predatory blue acaras (*Andinoacara pulcher*) reduce their maximum approach speed when repeatedly exposed to prey that escape unpredictably, enabling them to be as successful as those exposed to predictable prey. Under these two scenarios, if predators can learn to adjust their tactics to match the prey fluctuation regime they experience, both specialist and generalist hunters should achieve similar hunting success and coexist [@phillips.etal2017].

In this study, we use data from players in the online videogame *Dead by Daylight* to investigate 1) how repeated interactions with prey shape predator foraging specialization and 2) how the relationship between specialization and foraging success changes over time. *Dead by Daylight* simulates a predator-prey interaction across complex virtual environments, providing a system to test ecological hypotheses on foraging behaviour, group selection, and social interactions [reviewed in @montiglio2025]. In this game, four prey players forage for resources while avoiding predation by a fifth player. The predator population is composed of individuals that vary consistently in their use of hunting tactics along an ambush-cursorial continuum: at one extreme hunters move more slowly and ambush prey; at the other extreme, hunters roam the virtual environment at high velocities [@fraserfranco.etal2022]. To assess how foraging specialization emerges throughout experience, we define the level of individual foraging specialization as the intra-individual variance (IIV) in movement speed across matches following the definition by @cleasby.etal2015. Low IIV describes individual foraging specialization (i.e., always using an ambush or cursorial tactic) whereas high IIV describes a flexible use of foraging tactics (i.e., generalist predators switching between ambush and cursorial tactics, see Methods for details).

```{r fig1, out.width = "100%", fig.cap = "**Figure 1**. Hypothesis describing how prey-mediated individual foraging specialization emerges over repeated interactions. Panels A and B display two scenarios with different outcomes at the population level, but similar outcomes at the individual level. On panel A, all predators encounter prey with similar attributes (e.g., some encountered prey that were always fast while others encountered slower prey more often) through time. They differ in their mean behaviour (tactic) which is adjusted to the prey that they encountered. Because encounters where predictable, they all display similar degrees of individual foraging specialization, resulting in the population being specialized. On panel B, all predators encounter varying types of prey (e.g., in some encounters the prey were fast, in others the prey were slower), such that they also display similar degrees of individual foraging specialization as in panel A. However, in this case, they all become generalist hunters, resulting in the population distribution being wider. Panel C shows a different scenario, where predators display varying degrees of foraging specialization. In this case, some individuals consistently encountered similar prey, and thus specialized, while others encountered varying groups of prey, thus becoming generalists. Ultimately, those that learn to specialise on the proper tactic given prey encounters have greater success than those that didn't, resulting in no differences between specialists and generalists."}

@@

To test the hypothesis that repeated interactions with prey shape predator foraging specialization, we analyze the movement speed of predators and their prey over time. We expect predators to differ in the use of their hunting tactic (i.e., mean movement speed) which will be tuned to the mean speed of the prey that they encounter (Figure 1A-B). We predict that if all predators consistently encounter prey moving at similar speeds, then the population as a whole (i.e., all individuals) should specialize on moving at similar speeds, resulting in a narrow behavioural niche (Figure 1A). In contrast, if all predators encounter groups of prey moving at varying speeds, then they should all become more flexible in their tactic use (i.e., generalists), adjusting their speed to the prey from one encounter to the other, resulting in a higher population variance in foraging behaviour (Figure 1B). In both scenarios, differences among individuals in IIV across experience should be low (i.e., similar individual foraging specialization, Figure 1A-B). Alternatively, differences among individuals in foraging specialization may emerge if they experience different interactions with their prey. In this case, predators that encountered prey moving at similar speeds across experience should specialize on moving at a more fixed speed, while predators that encountered different prey speeds across matches should adopt a generalist movement strategy, resulting in an increase in among individual differences in IIV (Figure 1C). If we detect such prey-dependent fine-tuning with experience, then specialist and generalist hunters should attain equal success, resulting in no apparent relationship between foraging specialization and prey capture (Figure 1C).

@@

*DBD* is an asymmetric multiplayer survival game developed by Behaviour Interactive Inc. In this game, participants take on roles as either a predator or one of the prey, with each match consisting of one predator and four prey. A matchmaking system pairs groups of players with similar skill such that individuals should all have similar experience. The predator's primary goal is to hunt and capture the prey, while the prey aim to locate and collect resources in the form of power generators. When all generators are activated, the prey can escape through one of two exit gates and win the match. A match concludes when the predator eliminates all remaining prey that have not escaped or when the final prey successfully exits the virtual environment. Players, whether predator or prey, select avatars with distinct abilities tailored to specific playstyles (e.g. stealthy vs evasive prey, or stalking vs patrolling predators). During the study period, the game featured `r toString(format(length(unique(dat$avatar_id))))` predator avatars. The virtual environments in the game vary in size and structural complexity, combining fixed and procedurally generated elements such as vegetation, buildings, and maze-like structures. Troughout the study period, there were `r toString(format(length(unique(dat$environment_id))))` virtual game environments available for gameplay.

@@

The videogame company provided six months of gameplay data, encompassing all player activity from `r toString(format(min(dat$date)))` to `r toString(format(max(dat$date)))`. We focused exclusively on matches played in "Online" mode, where players were strangers to one another. We excluded matches where players were inactive, as indicated by mean movement speeds that were zero or nearly zero. Additionally, matches where players were suspected of hacking or engaging in unintended gameplay behaviour were also removed based on our knowledge of the game mechanics. We then selected a cohort of players that played 300 matches or more, and monitored all their matches from the first to a maximum of 500 matches.

Our dataset includes `r toString(format(length(unique(dat$predator_id))))` unique predator players, collectively accounting for `r toString(formatC(length(unique(dat$match_id)), big.mark = " "))` matches. The number of matches played by predator players ranged from `r toString(format(min(dat$total_xp_pred)))` to `r toString(format(max(dat$total_xp_pred)))`. Match durations varied between `r toString(format(round(min(dat$game_duration / 60))), digits = 2)` and `r toString(format(round(max(dat$game_duration / 60))), digits = 2)` minutes, with an average duration of `r toString(format(round(mean(dat$game_duration / 60))), digits = 2)` minutes. The following details were recorded for each match: the anonymous player ID, the chosen predator avatar, the game environment, the predator's level of experience, their speed, the average speed of the prey group they encountered, and the average rank of those prey (used as a proxy for prey skill). The game's ranking system, designed to match players of similar skill levels (https://deadbydaylight.fandom.com/wiki/Rank), was incorporated into our analysis to detect any relationship between predator experience and foraging success [see @fraserfranco.etal2024].

We analyzed the predator's mean speed and the mean speed of the prey group encountered by the predator. The predator's mean speed is measured as the mean distance traveled per second during a match (mean= `r toString(format(round(mean(dat$pred_speed), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$pred_speed), digits = 2), nsmall = 2))` m/s). We measured the preys' speed as the mean travel speed of the four individual prey in a match (mean = `r toString(format(round(mean(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` m/s). We defined hunting success as the number of prey consumed during the match (min = `r toString(format(min(dat$hunting_success)))`, max = `r toString(format(max(dat$hunting_success)))`). Lastly, we categorized predators for each match based on their cumulated experience [see @fraserfranco.etal2024]. We labeled predators as novices for matches where they had cumulated less than 100 matches, intermediate for matches where they had cumulated between 100 and 299 matches, and advanced for matches where they had cumulated more than 299 matches (max 500). Since our goal was to monitor predator players throughout their experience and that they all played at least 300 matches, they all appeared in each of the three experience categories.

@@

We tested whether foraging behaviour changed with experience, and compared the success of predators along the flexible-specialist hunter continuum, using a multivariate double-hierarchical generalized linear model (MDHGLM) [@lee.nelder2006;@cleasby.etal2015;@mitchell.etal2016]. For each experience level, we fitted the model on three traits: the predator's speed ($Y_{1}$) and the prey's speed ($Y_{2}$) using a Gaussian error distribution, and the predator's hunting success ($Y_{3}$) using a beta-binomial error distribution. Fixed and random effect structure were trait-specific and the same for all experience levels. The predator and the prey speed included a mean and a dispersion part (eqn 1.1, 1.2 and 2.1, 2.2). For both traits, we controlled for the standardized prey rank on both the mean and the dispersion parts of the model. We included game environment, predator avatar, and individual predator identity random effects on the mean part of the model. The dispersion part included individual predator identity random effects exclusively. For hunting success, we modeled the mean part only (eqn 3) and controlled for the standardized prey rank and game duration, and included an individual identity random effect.

The model simultaneously estimated five among-individual variance components with their covariances for each experience level, resulting in a 15x15 (co)variance matrix. For each experience level, the model estimated among individual variance in mean predator speed, mean prey speed, and mean hunting success. This allows us to test for individual differences in tactic use (i.e., along the slow-fast continuum), in the mean speed of the prey encountered, and in mean prey consumption. The model also estimated among individual variance in intra-individual variation (IIV) for both prey and predator speed. This allows us to test if predators do not experience the same degree of prey heterogeneity and if they differ in individual specialization. Finally, the model estimated all pairwise covariances at the individual level among traits, among experience, and across traits and experience. The resulting 15x15 (co)variance matrix thus provides the structure of the predator-prey trait interactions as well as the relationship between specialization and success across experience at the individual-level.

@@

We performed our analyses under a Bayesian framework. We used weakly informative Gaussian priors for the prey rank ($N(0,\ 1^{2})$: eqn. 1.1 to 2.2) and game duration ($N(0.5,\ 0.5^{2})$: eqn. 3). Based on previous results showing that the mean predator speed revolves around 3 m/s [@fraserfranco.etal2022], we applied a weakly informative Gaussian prior on the intercept for predator and prey speed ($N(3,\ 0.5^{2})$: eqn. 1.1 and 2.1). Similarly, we used a weakly informative Gaussian prior on the intercept for the dispersion part ($N(0,\ 1^{2})$: eqn 1.2 and 2.2). For hunting success, we applied a positive Gaussian prior on the precision parameter ($N(2,\ 1^{2})$: eqn. 3) and a negative Gaussian prior ($N(-4,\ 1^{2})$: eqn 3) on the intercept because hunting success should be close to 0 at the lowest game duration. For every predicted variable, we applied a weakly informative half-Gaussian prior ($N(0,\ 1^{2})$) on the random effects for the mean part and the dispersion part of the model. We used a regularizing $LKJ(2)$ prior on the correlation matrix.

@@

We parametrized the MDHGLM to run 2500 iterations with a thinning set to eight, with the first 500 used as warm up iterations (yielding 1000 posterior samples for each parameter). We assessed the convergence of the chains using trace plots, R-hat diagnostics with a threshold of <1.01, and effective sample sizes (ESS) with a threshold of >100 [@vehtari.etal2021]. We also performed posterior predictive checks which showed an adequate fit of the models (for details, see https://github.com/quantitative-ecologist/experience-hunting-tactics). We report all the posterior parameter estimates using the median of the posterior distribution with the highest posterior density (HPD) intervals at 50%, 80%, and 95%.

@@

The mean speed of the predator population remained stable with experience (Figure 2 and Appendix 1: Table S1). Likewise, the variation in speed did not change with experience, indicating that foraging specialization remained stable at the population level (Figure 2 and Appendix 1: Table S1). At the individual level, predators differed slightly in their mean speed across experience levels, indicating marginal differences in hunting tactics (Figure 2 and Appendix 1: Table S2). However, they displayed important differences in foraging specialization (Appendix 1: Table S2) which also increased slightly with experience as there was a 0.11 unit increase in among individual differences in IIV from novice to advanced (Figure 1). `r toString(scales::percent(percent_tab2[1,]$percentage, accuracy = 1))` of the population switched from a flexible to a specialized hunting tactic, and vice-versa, as predators gained experience (i.e., >0.2 change in standard deviation; Figure 2). In contrast, `r toString(scales::percent(percent_tab2[2,]$percentage, accuracy = 1))` displayed lower changes (i.e., >0.05 and <0.2 change in standard deviation) and `r toString(scales::percent(percent_tab2[3,]$percentage, accuracy = 1))` showed almost no change with experience (i.e., <0.05 change in standard deviation).

```{r fig2, out.width = "100%", fig.cap = "**Figure 2**. Median posterior differences and HPD intervals comparing predator foraging behaviour at the population and individual level across experience stages. The behavioural parameter predicted by the MDHGLM being compared is on the y axis, and the difference in parameter values between two experience stages is on the x axis. The pairwise comparisons between experience stages are displayed across the three panels A, B, and C. We compare the mean speed of the predator population, the speed of the prey it encountered, and its hunting success across experience (population mean). We also compare the population's foraging specialization and variation in speed of the prey it encountered across experience (population variance). At the individual level, we compare among individual differences in speed, the speed of the prey encountered, and hunting success (individual variation mean). Lastly, we compare individual variation in IIV across experience, indicating whether individual differences in foraging specialization and in the variation of the groups of prey encountered changed with experience (individual variation IIV)."}

@@

As the predator population gained experience, there was strong evidence that the prey increased their speed, although slightly (Figure 1 and Appendix 1: Table S1). The population also encountered groups of prey using different speeds through time (Figure 1 and Appendix 1: Table S1). At the individual level, all predators encountered prey with similar average speeds throughout experience (Figure 1 and Appendix 1: Table S2). Yet, even if they experienced similar average prey speeds across all groups encountered, there were important differences among predators in how (dis)similar were the groups they encountered (Appendix 1: Table S2). Thus, some predators encountered similar groups (i.e., all groups using similar speeds) while others encountered distinct groups (i.e., some slower and some faster). These differences among individuals in IIV for prey encounters increased only marginally with experience (Figure 1).

The predator population's mean hunting success increased with experience (Figure 1 and Appendix 1: Table S1), with mean prey consumption revolving around two prey. There were also important differences among individuals in mean hunting success (Appendix 1: Table S2), but they remained stable across experience (Figure 1).

```{r fig3, out.width = "100%", fig.cap = "**Figure 3**. Individual behavioural distribution of the predators' speed drawn from the MDHGLM estimates for individuals that had (A) the greatest increase in specialization and (B) the greatest increase in flexibility. The background (gray) distributions are for when individuals were novice hunters, and the foreground (blue) distributions with solid contour lines are for when they were advanced hunters. Both figure panels are ordered by ascending degree of increase in either specialization or flexibility. We built the figure by first subtracting the estimated standard deviation of all individuals as novices with their standard deviation as advanced hunters, and then selected those with the greatest difference using the 25% and 75% quantiles. Individuals on panel A have an increase in specialization equal or greater than 0.2 standard deviations, while those on panel B have an increase in flexibility equal to or greater than 0.2 standard deviations."}

@@

Predators that were on average faster had lower IIV in speed (Figure 3A-B). Thus, faster predators specialized on fast movement, while slower predators were more flexible in their movement. As predators gained experience, these strategies were increasingly defined as the correlation changed by 0.16 units, from -0.62 (-0.70, -0.55 $95\%\ CI$) for when predators were novices to -0.78 (-0.83, -0.72 $95\%\ CI$) as they reached the advanced stage (Figure 3C). Predators that were on average faster also tended to encounter groups of prey that were on average faster and more similar (Figure 3A-B). There was however no evidence that these correlations increased with experience (Figure 3C)

>>>>

---
title: "Individual foraging specialization and success change across experience in a virtual predator-prey system"
output: 
  word_document:
    reference_docx: manuscript-styles.docx
    fig_caption: yes
bibliography: references.bib
csl: ecology.csl
---

@@

library(data.table)
library(brms)
library(flextable)
library(officer)
library(dplyr)
library(equatags)

@@

The capacity of predators to match their hunting tactics to their prey and to optimize their skills at using them over successive encounters is hypothesized to promote individual foraging specialisation. Through this process, predators should be able to capture more prey. Yet, there are limited empirical assessments showing links between past experience, foraging specialization, and hunting success at the individual level, due to the challenges of monitoring direct interactions in the wild. In this study, we used a virtual predator-prey system (the game *Dead by Daylight*) to investigate how individual predator foraging specialization and success developed across repeated interactions with prey. We found that 68% of predators became either increasingly specialized by always moving at a fast pace (specialised cursorial tactic), or generalists by transitioning between slow and fast speeds (generalist ambush tactic). The predators' strategies were partially matched to their prey's speed, suggesting that changes in hunting behaviour were driven by repeated encounters with their prey. Specialist and generalist foragers achieved similar success overall, although specialist hunters had a slight advantage. Hence, our findings suggest that experience may promote behavioural diversification in predator-prey systems.

@@

Recent evidence suggests that the resource fluctuations predators experience over time (e.g., temporal changes in prey behaviour) may be key in shaping individual behavioural specialization and its fitness consequences [@patrick.weimerskirch2014b;@vandenbosch.etal2019;@santoro.etal2019;@manlick.etal2021]. Theory outlines two contrasting scenarios, with outcomes defined by the predator's ability to learn the optimal strategy in response to resource variability. In the first scenario, individuals should gradually specialize in a foraging tactic (e.g., always moving fast) if it enables them to consistently achieve successful prey captures while learning to hunt in environments with predictable (or stable) resources [@tinker.etal2009;@dukas2019]. Such conditions favor specialist hunters, as they require less energy and time to search for, capture, and handle prey [@weimerskirch2007;@woo.etal2008;@potier.etal2015]. For example, habitat specialization under stable resources is associated with reduced foraging effort and increased offspring growth rates in Herring Gulls (*Larus argentatus*) [@vandenbosch.etal2019]. In the second scenario, predators are expected to be more generalist by learning to use a variety of hunting tactics when they hunt in highly variable environments. Indeed, while there can be costs to attempt novel tactics by trial and error, predators can reduce the consequences of uncertainty in environments under fluctuating resources by learning to adjust their tactic to the prey that they encounter [@dall2010;@mathot.etal2012;@stephens1993;@ishii.shimada2010;@kelley.magurran2011]. @szopa-comley.ioannou2022 showed that predatory blue acaras (*Andinoacara pulcher*) reduce their maximum approach speed when repeatedly exposed to prey that escape unpredictably, enabling them to be as successful as those exposed to predictable prey. Under these two scenarios, if predators can learn to adjust their tactics to match the prey fluctuation regime they experience, both specialist and generalist hunters should achieve similar hunting success and coexist [@phillips.etal2017].

In this study, we use data from players in the online videogame *Dead by Daylight* to investigate 1) how repeated interactions with prey shape predator foraging specialization and 2) how the relationship between specialization and foraging success changes over time. *Dead by Daylight* simulates a predator-prey interaction across complex virtual environments, providing a system to test ecological hypotheses on foraging behaviour, group selection, and social interactions [reviewed in @montiglio2025]. In this game, four prey players forage for resources while avoiding predation by a fifth player. The predator population is composed of individuals that vary consistently in their use of hunting tactics along an ambush-cursorial continuum: at one extreme hunters move more slowly and ambush prey; at the other extreme, hunters roam the virtual environment at high velocities [@fraserfranco.etal2022].

```{r fig1, out.width = "100%", fig.cap = "**Figure 1**. Hypothesis describing how prey-mediated individual foraging specialization emerges over repeated interactions. Panels A and B display two scenarios with different outcomes at the population level, but similar outcomes at the individual level. A) all predators encounter prey with similar attributes (e.g., some encountered prey that were always fast while others encountered slower prey more often) through time. They differ in their mean behaviour (tactic) which is adjusted to the prey that they encountered. Because encounters where predictable, they all display similar degrees of individual foraging specialization, resulting in the population being specialized. B) all predators encounter varying types of prey (e.g., in some encounters the prey were fast, in others the prey were slower), such that they also display similar degrees of individual foraging specialization as in panel A. However, in this case, they all become generalist hunters, resulting in the population distribution being wider. C) A scenario where predators display varying degrees of foraging specialization. In this case, some individuals consistently encountered similar prey, and thus specialized, while others encountered varying groups of prey, thus becoming generalists. Ultimately, those that learn to specialise on the proper tactic given prey encounters have greater success than those that didn't, resulting in no differences between specialists and generalists."}

@@

To test the hypothesis that repeated interactions with prey shape predator foraging specialization, we analyze the movement speed of predators and their prey over time. To assess how foraging specialization emerges throughout experience, we define the level of individual foraging specialization as the intra-individual variance (IIV) in movement speed across matches following the definition by @cleasby.etal2015. Low IIV describes individual foraging specialization (i.e., always using an ambush or cursorial tactic) whereas high IIV describes a flexible use of foraging tactics (i.e., generalist predators switching between ambush and cursorial tactics, see Methods for details). First, we expect predators to differ in the use of their hunting tactic (i.e., mean movement speed) which will be tuned to the mean speed of the prey that they encounter (Figure 1A-B). Second, we predict that if all predators consistently encounter prey moving at similar speeds, then the population as a whole (i.e., all individuals) should specialize on moving at similar speeds, resulting in a narrow behavioural niche (Figure 1A). In contrast, if all predators encounter groups of prey moving at varying speeds, then they should all become more flexible in their tactic use (i.e., generalists), adjusting their speed to the prey from one encounter to the other, resulting in a higher population variance in foraging behaviour (Figure 1B). In both scenarios, differences among individuals in IIV across experience should be low (i.e., similar individual foraging specialization, Figure 1A-B). Alternatively, differences among individuals in foraging specialization may emerge if they experience different interactions with their prey. In this case, predators that encountered prey moving at similar speeds across experience should specialize on moving at a more fixed speed, while predators that encountered different prey speeds across matches should adopt a generalist movement strategy, resulting in an increase in among individual differences in IIV (Figure 1C). Lastly, if we detect such prey-dependent fine-tuning with experience, then specialist and generalist hunters should attain equal success, resulting in no apparent relationship between foraging specialization and prey capture (Figure 1C).

@@

*DBD* is an asymmetric multiplayer survival game developed by Behaviour Interactive Inc. In this game, participants take on roles as either a predator or one of the prey, with each match consisting of one predator and four prey. A matchmaking system pairs groups of players with similar skill levels such that individuals should all have similar profiency. The predator's primary goal is to hunt and capture the prey while the prey must locate and collect resources in the form of power generators that must be activated. When all generators are activated, the prey can escape through one of two exit gates and win the match. A match ends when the predator has eliminated all remaining prey that have not escaped or when the last prey successfully exits the virtual environment. Players, whether predator or prey, select avatars with distinct abilities designed for specific playstyles (e.g. stealthy vs evasive prey, or stalking vs patrolling predators). During the study period, the game featured `r toString(format(length(unique(dat$avatar_id))))` predator avatars. The virtual environments in which matches are played vary in size and structural complexity, combining fixed and procedurally generated elements such as vegetation, buildings, and maze-like structures. Troughout the study period, there were `r toString(format(length(unique(dat$environment_id))))` virtual game environments available for gameplay.

@@

The dataset comprised gameplay data from `r toString(format(min(dat$date)))` to `r toString(format(max(dat$date)))`. We focused exclusively on matches played in "Online" mode, where players were strangers to one another and paired by a matchmaking algorithm. We excluded matches where players were inactive, as indicated by null or nearly null movement speeds, and matches where players were suspected of hacking or engaging in unintended gameplay behaviour based on our knowledge of the game mechanics. We then selected a cohort of players that played 300 matches or more, and monitored all their matches from the first to a maximum of 500 matches.

Our dataset includes `r toString(format(length(unique(dat$predator_id))))` unique predator players, collectively contributing to `r toString(formatC(length(unique(dat$match_id)), big.mark = " "))` matches. The number of matches played by predator players ranged from `r toString(format(min(dat$total_xp_pred)))` to `r toString(format(max(dat$total_xp_pred)))`. Match durations varied between `r toString(format(round(min(dat$game_duration / 60))), digits = 2)` and `r toString(format(round(max(dat$game_duration / 60))), digits = 2)` minutes, with an average duration of `r toString(format(round(mean(dat$game_duration / 60))), digits = 2)` minutes. The final dataset provided the following variables: the player ID, the chosen predator avatar, the game environment, the predator's level of experience, the mean movement speed of the predator, the mean movement speed of the prey group, and their average rank (used as a proxy for prey skill). The game's ranking system, designed to match players of similar skill levels (https://deadbydaylight.fandom.com/wiki/Rank), was incorporated into our analysis to detect any relationship between predator experience and foraging success [see @fraserfranco.etal2024].

We analyzed the predator's mean speed and the mean speed of the prey group encountered by the predator. The predator's mean speed is measured as the mean distance traveled per second during a match (mean= `r toString(format(round(mean(dat$pred_speed), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$pred_speed), digits = 2), nsmall = 2))` m/s). We measured the preys' speed as the mean travel speed of the four individual prey in a match (mean = `r toString(format(round(mean(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` m/s). We defined hunting success as the number of prey consumed during the match (min = `r toString(format(min(dat$hunting_success)))`, max = `r toString(format(max(dat$hunting_success)))`). Lastly, we categorized predators for each match based on the experience they cumulated [see @fraserfranco.etal2024]. We labeled predators as novices when they played between 1 and 99 matches, as intermediates between 100 and 299 matches, and as advanced between 300 and 500 matches.

@@

 We used a multivariate double-hierarchical generalized linear model (MDHGLM) to quantify a) changes in the average foraging tactic (i.e., the mean movement speed) and in the degree of specialization (i.e., IIV in movement speed) over successive matches for each player; and b) the relationship between the average tactic and specialization, and hunting success [@lee.nelder2006;@cleasby.etal2015;@mitchell.etal2016]. For each experience level, we fitted the model on three traits: the predator's speed ($Y_{1}$) and the prey's speed ($Y_{2}$) using a Gaussian error distribution, and the predator's hunting success ($Y_{3}$) using a beta-binomial error distribution. Fixed and random effect structure was trait-specific and the same for all experience levels. For the predator and the prey speed, we modeled the mean and the residual standard deviation (eqn 1.1, 1.2 and 2.1, 2.2). For both traits, we controlled for the standardized prey rank on the mean and the residual standard deviation. We included game environment, predator avatar, and individual predator identity random effects on the mean part of the model. The residual standard deviation included individual predator identity random effects exclusively. For hunting success, we modeled the mean only (eqn 3) and controlled for the standardized prey rank and game duration, and included an individual identity random effect.

The model simultaneously estimated five among-individual variance components with their covariances for each experience level, resulting in a 15x15 (co)variance matrix. For each experience level, the model estimated among individual variance in mean predator speed, mean prey speed, and mean hunting success. This allows us to test for individual differences in tactic use (i.e., along the ambush-cursorial continuum), in the mean speed of the prey encountered, and in mean prey consumption. The model also estimated among individual variance in intra-individual variation (IIV) for both prey and predator speed. This allows us to test if predators do not experience the same degree of prey heterogeneity and if they differ in individual specialization. Finally, the model estimated all pairwise covariances at the individual level among traits, among experience, and across traits and experience. The resulting 15x15 (co)variance matrix thus provides the structure of the predator-prey trait interactions as well as the relationship between specialization and success across experience at the individual-level.

@@

We performed our analyses within a Bayesian framework. We used weakly informative Gaussian priors for the prey rank ($N(0,\ 1^{2})$: eqn. 1.1 to 2.2) and game duration ($N(0.5,\ 0.5^{2})$: eqn. 3). Based on previous results showing that the mean predator speed revolves around 3 m/s [@fraserfranco.etal2022], we applied a weakly informative Gaussian prior on the intercept for predator and prey speed ($N(3,\ 0.5^{2})$: eqn. 1.1 and 2.1). Similarly, we used a weakly informative Gaussian prior on the intercept for the dispersion part ($N(0,\ 1^{2})$: eqn 1.2 and 2.2). For hunting success, we applied a positive Gaussian prior on the precision parameter ($N(2,\ 1^{2})$: eqn. 3) and a negative Gaussian prior ($N(-4,\ 1^{2})$: eqn 3) on the intercept because hunting success should be close to 0 at the lowest game duration. For every predicted variable, we applied a weakly informative half-Gaussian prior ($N(0,\ 1^{2})$) on the random effects for the mean part and the dispersion part of the model. We used a regularizing $LKJ(2)$ prior on the correlation matrix.

@@

We parametrized the MDHGLM to run 2500 iterations on four chains with a thinning interval set at eight iterations, with the first 500 used as warm up iterations (yielding 1000 posterior samples for each parameter). We assessed the convergence of the chains using trace plots, R-hat diagnostics with a threshold of <1.01, and effective sample sizes (ESS) with a threshold of >100 [@vehtari.etal2021]. We also performed posterior predictive checks which showed an adequate fit of the models (for details, see https://github.com/quantitative-ecologist/experience-hunting-tactics). We report all the posterior parameter estimates using the median of the posterior distribution with the highest posterior density (HPD) intervals at 50%, 80%, and 95%.

@@

At the population level, the mean predator speed remained stable as predators progressed from novices to advanced hunters (Figure 2A and Appendix 1: Table S1). Likewise, the variation in speed did not change with experience, indicating that foraging specialization remained stable at the population level (Figure 2A and Appendix 1: Table S1). At the individual level, differences among predators in their mean speed were weak and did not change across experience (Figure 2A and Appendix 1: Table S2). However, predators displayed minor differences in foraging specialization as novices, with differences increasing slightly as individuals gained experience (Figure 2A and Appendix 1: Table S2). `r toString(scales::percent(percent_tab2[1,]$percentage, accuracy = 1))` of the population switched from a flexible to a more specialized hunting tactic, and vice-versa, as predators gained experience (i.e., >0.2 change in standard deviation; Figure 3). In contrast, `r toString(scales::percent(percent_tab2[2,]$percentage, accuracy = 1))` displayed lower changes (i.e., >0.05 and <0.2 change in standard deviation) and `r toString(scales::percent(percent_tab2[3,]$percentage, accuracy = 1))` showed almost no change with experience (i.e., <0.05 change in standard deviation).

```{r fig2, out.width = "100%", fig.cap = "**Figure 2**. Median posterior differences and HPD intervals comparing predator foraging behaviour, prey behaviour, and predator hunting success at the population and individual level across experience stages. The parameters being compared are on the y axis, and the differences in parameter values between two experience stages are on the x axis. The pairwise comparisons between experience stages are displayed as different point shapes on each panel. A) For the predator speed, we compare the foraging tactic (mean) and specialisation (variance) at the population level and at the individual level between each experience stage. B) For the prey speed, we compare the mean and the variance in speed of prey encountered at the population and individual level between each experience stage. C) For the predator hunting success, we compare the mean success at the population and individual level between each experience stage."}

@@

At the population level, we found strong evidence that the prey increased their speed slightly as predators gained experience (Figure 2B and Appendix 1: Table S1). The predators also encountered a greater variation of prey speeds at later experience stages (Figure 2B and Appendix 1: Table S1). At the individual level, predators encountered prey with similar average speeds across experience stages (Figure 2B and Appendix 1: Table S2). However, predators differed in the range of prey speeds they encountered and the magnitude of these differences increased as they gained experience (Figure 2B and Appendix 1: Table S2). Thus, some predators encountered prey moving at similar speeds from one encounter to the next (i.e., all groups using similar speeds) while others encountered prey moving at different speeds (i.e., some slower and some faster).

The predator population's mean hunting success increased with experience (Figure 2C and Appendix 1: Table S1), with mean prey consumption reaching around two prey. We did not find evidence for differences among individuals in mean hunting success at any experience stage, even though differences increased slightly (Figure 2C and Appendix 1: Table S2).

```{r fig3, out.width = "100%", fig.cap = "**Figure 3**. Individual behavioural distribution of the predators' speed drawn from the MDHGLM estimates for individuals that had (A) the greatest increase and (B) the greatest decrease in specialisation. The background (gray) distributions are for when individuals were novice hunters, and the foreground (blue) distributions with solid contour lines are for when they were advanced hunters. Both figure panels are ordered by ascending degree of change in specialization. We built the figure by first subtracting the estimated standard deviation of all individuals as novices with their standard deviation as advanced hunters, and then selected those with the greatest difference using the 25% and 75% quantiles. Individuals on panel A have an increase in specialization equal or greater than 0.2 standard deviations, while those on panel B have an increase in flexibility equal to or greater than 0.2 standard deviations."}

@@

Predators that were on average faster had lower IIV in speed (Figure 4A-B). Thus, faster predators specialized to a higher degree than slower predators. As predators gained experience, these strategies were increasingly defined as the correlation changed from -0.62 (-0.70, -0.55 $95\%\ CI$) among novices to -0.78 (-0.83, -0.72 $95\%\ CI$) among advanced hunters (Figure 4C). Faster predators tended to encounter faster prey; they also encountered groups of prey groups of prey that were more similar from match to match (Figure 4A-B), irrespective of their experience (Figure 4C).

There was no correlation (-0.02, -0.14, 0.01 $95\%\ CI$) between mean prey speed and IIV in predator speed when predators were novice (Figure 4A), but the correlation increased to -0.16 (-0.27, -0.04 $95\%\ CI$) at the advanced stage (Figure 4B), such that advanced hunters that encountered faster prey tended to be more specialized (Figure 4C). Predators with a larger IIV in speed encountered a greater range of prey speeds, but this effect was weak (Figure 4A-B). Hence, predators that encountered groups of prey with similar speeds tended to be more specialized in their tactic, while those that encountered groups of prey with varying speeds were less specialised. However, this correlation did not change with experience (Figure 4C).
